\normalfont \LARGE Improved Quasi-Static Methods for Time-Dependent Neutron Diffusion and Implementation in Rattlesnake

### 2016-12-20 doc/Thesis_proposal/thesis_proposal.tex
final comments on the MS proposal

Transient modeling of nuclear reactors has been a chronically foreboding task due to its computationally expensive nature.  However, recent developments in nuclear testing, including the revitalization of the Transient Reactor Testing (TREAT) Facility at Idaho National laboratory (INL), have brought significant attention to the development of transient reactor modeling.  Transient computational methods currently in use at national laboratories, including INL, have proven to be overbearingly computationally expensive for complex, real-world scenarios.  Therefore, implementation of more efficient and faster methods is highly desired.  The Department of Energy (DOE) has especially sought to develop in the MOOSE framework.  \\

MOOSE is a multiphysics framework being developed at INL that presents the architecture for physics-based applications \cite{moose}.  Rattlesnake is a large application in MOOSE that involves deterministic radiation transport physics.  Rattlesnake is able to solve steady-state, transient, and k-eigenvalue neutron transport and diffusion problems using finite element method (FEM) \cite{wang2013}. \\

The difficulty in transient reactor modeling is due to the neutron transport equation, which describes the flux of neutrons. The difficulty in solving or evaluating the equation comes from the dimensionality of the phase-space and the fact that the time discretization has to be implicit, which  leads to kinear system solves.  The equation has seven independent variables: space ($\vec{r}$), energy ($E$), direction ($\vec{\Omega}$), and time ($t$) \cite{duderstadt1976nuclear}.  Nuclear reactor simulations often utilize the neutron diffusion approximation, which carefully eliminates the dependence on direction.  For the purposes of this research, neutron diffusion will be discussed exclusively. \\

For computational purposes, the neutron diffusion equation is discretized in space, energy, and time.  There are several viable discretization schemes for each of these variables, but FEM for space and multigroup method for energy were used for the development of this research. Since the purpose of this research involves temporal dependence, discretization in time was kept general.  The time dependent neutron diffusion equation with delayed neutron precursors can be seen in Equations~\ref{eq:flux}~and~\ref{eq:precursor}.

\begin{subequations}
\begin{align}
\frac{1}{v^g}\frac{\partial \phi^g}{\partial t}  =& \frac{\chi_p^g}{\keff} (1-\beta)\sum_{g'=1}^G  \nu^{g'} \Sigma_f^{g'} \phi^{g'} -  \left( -\div D^g \grad  + \Sigma_r^g \right) \phi^g  \nonumber \\
&  + \sum_{g'\neq g}^G\Sigma_s^{g'\to g} \phi^{g'}  + \sum_{i=1}^I\chi_{d,i}^g\lambda_i C_i \ , \quad 1 \le g \le G 
\label{eq:flux}
\end{align}
\be
\frac{dC_i}{dt} = \frac{\beta_i}{\keff}\sum_{g=1}^G\nu^{g} \Sigma_f^g \phi^{g} - \lambda_i C_i \ , \quad 1 \le i \le I 
\label{eq:precursor}
\ee
\end{subequations}

@@

Most reactor computation frameworks, including Rattlesnake, discretize the time variable directly with these equations using a multitude of schemes (Implicit Euler, Crank-Nicholson, Runge-Kutta, etc.).  In this paper, this method of discretizing Equations~\ref{eq:flux}~and~\ref{eq:precursor} is considered implicit discretization.  This research intends to improve upon this method by instead implementing the improved quasi-static method (IQS). \\

@@

These techniques are by no means an exhaustive list of the possible iteration techniques for IQS. Dulla et al. in \cite{Dulla2008} does an in depth analysis of the fixed point iteration technique most similar to Sissaoui and Koclas, involving convergence rates and solution results.  However, no comprehensive analysis of iteration techniques exists.

@@

\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
\item Investigate iteration convergence for previously implemented iterative schemes and newly conceived one described in \sct{sect:inv_iter}.
\item Validate and verify IQS method with several low and high order time discretization schemes.
\item Implement step doubling time adaptation with IQS and analyze its performance compared to implicit discretization.
\item Apply temperature feedback with IQS and analyze its performance with multiphysics benchmarks.
\item Implement IQS into the Rattlesnake framework at INL.
\item Perform testing with IQS to benchmarks including TWIGL, LRA, and C5G7-2D.
\end{enumerate}

>>>>

Transient modeling of nuclear reactors has been a chronically foreboding task due to its computationally expensive nature.  However, recent developments in nuclear testing, including the revitalization of the Transient Reactor Testing (TREAT) Facility at Idaho National laboratory (INL), have brought significant attention to the development of transient reactor modeling.  3-D transient computational methods currently in use at national laboratories, including INL, have proven to be overbearingly computationally expensive for complex, real-world scenarios.  Therefore, implementation of more efficient and faster methods for 3-D transient simulations is highly desired.  The Department of Energy (DOE), through its NEAMS (Nuclear Energy Advanced Modeling and Simulation) program, has especially sought the development of transient multiphysics capability in 
INL's MOOSE framework.  \\

MOOSE is a multiphysics framework being developed at INL that presents the architecture for physics-based applications \cite{moose}.  Rattlesnake is a large application in MOOSE that involves deterministic radiation transport physics.  Currently, Rattlesnake is able to solve steady-state, transient, and k-eigenvalue neutron transport and diffusion problems using finite element methods (FEM) \cite{wang2013}. \\

The difficulty in transient reactor modeling is due to 
\begin{enumerate}
\item
the high-dimensionality of the phase-space for the governing equations that describe the flux of neutrons (6-D+time for multigroup neutron transport and 4-D+time for multigroup neutron diffusion), and 
\item
the fact that the time discretization has to be implicit, which  leads to stiff system of equations.  
\end{enumerate}
The transport equation has seven independent variables: space ($\vec{r}$), energy ($E$), direction ($\vec{\Omega}$), and time ($t$) \cite{duderstadt1976nuclear}.  Nuclear reactor simulations often utilize the neutron diffusion approximation, which carefully eliminates the dependence on direction.  For the purpose of this research, neutron diffusion will be discussed exclusively. \\

For computational purposes, the neutron diffusion equation is discretized in space, energy, and time.  There are several viable discretization schemes for each of these variables; FEM for space and multigroup in energy are used for the development of this research. Since the purpose of this research involves temporal dependence, discretization in time was kept general.  The time dependent neutron diffusion equation with delayed neutron precursors can be seen in Equations~\ref{eq:flux}~and~\ref{eq:precursor}.

\begin{subequations}
\begin{multline}
\frac{1}{v^g}\frac{\partial \phi^g}{\partial t}  = 
\frac{\chi_p^g}{\keff} (1-\beta)\sum_{g'=1}^G  \nu^{g'} \Sigma_f^{g'} \phi^{g'} 
+ \sum_{g'\neq g}^G\Sigma_s^{g'\to g} \phi^{g'}  \\ + \sum_{i=1}^I\chi_{d,i}^g\lambda_i C_i 
-  \left( -\div D^g \grad  + \Sigma_r^g \right) \phi^g   
\ , \quad 1 \le g \le G 
\label{eq:flux}
\end{multline}
\be
\frac{dC_i}{dt} = \frac{\beta_i}{\keff}\sum_{g=1}^G\nu^{g} \Sigma_f^g \phi^{g} - \lambda_i C_i \ , \quad 1 \le i \le I 
\label{eq:precursor}
\ee
\end{subequations}

@@

Most reactor computation frameworks, including Rattlesnake, discretize the time variable directly with these equations using a multitude of schemes (Implicit Euler, Crank-Nicholson, implicit Runge-Kutta, etc.).  In this paper, this method of discretizing Equations~\ref{eq:flux}~and~\ref{eq:precursor} is generally referred to as ``implicit discretization''.  This research intends to improve upon this method by instead implementing the improved quasi-static method (IQS) for neutron kinetics and
implement it within a multiphysics setting. \\

@@

These techniques are by no means an exhaustive list of the possible iteration techniques for IQS. Dulla et al. in \cite{Dulla2008} does an in depth analysis of the fixed point iteration technique most similar to Sissaoui and Koclas, involving convergence rates and solution results.  However, no comprehensive analysis of iteration techniques exists.
\tcr{last sentence is a bit vague}

@@

\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
\item Investigate iteration convergence for previously implemented iterative schemes and newly conceived one described in \sct{sect:inv_iter}.
\item Verify IQS method with several low and high order time discretization schemes.
\item Implement step doubling time adaptation with IQS and analyze its performance compared to implicit discretization of the flux equations.
\item Apply temperature feedback with IQS and analyze its performance with multiphysics benchmarks.
\item Implement IQS into the Rattlesnake framework at INL.
\item Perform testing with IQS to benchmarks including TWIGL and LRA. \tcr{what about the LMR benchmark? needed for paper}
\end{enumerate}
